---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---


## Loading and preprocessing the data

First we load appropriate libraries, set some options, then unzip and load the data:

```{r, echo=TRUE}
suppressWarnings(suppressMessages(library(dplyr)))
suppressWarnings(suppressMessages(library(ggplot2)))
options(scipen=999) # Get numbers to print nicely...

unzip('activity.zip')
data = read.csv('activity.csv', stringsAsFactors = F)

```

## What is mean total number of steps taken per day?

The following code computes total steps taken per day:

```{r, echo=TRUE}
stepsPerDay = data %>% # Start with the data we loaded
  filter(!is.na(steps)) %>% # Remove entries with NA steps
  group_by(date) %>% # Group into days
  summarise(totalSteps = sum(steps))  #Find total steps each day
```

Then we plot a histogram.

```{r, echo=T, fig.height=6, fig.width=8}

ggplot() + geom_histogram(data = stepsPerDay, aes(x = totalSteps, fill = 'coral'), binwidth = 1000 ) + theme(legend.position="none")

```

```{r, echo=TRUE}
meanStepsPerDay = mean(stepsPerDay$totalSteps)
print(meanStepsPerDay)
medianStepsPerDay = median(stepsPerDay$totalSteps)
```

So the mean total number of steps taken per day is `r meanStepsPerDay`, and the median is `r medianStepsPerDay`. 

## What is the average daily activity pattern?

The following code computes average steps taken in each 5 minute interval:

```{r, echo=TRUE}
stepsPerInterval = data %>% # Start with the data we loaded
  filter(!is.na(steps)) %>% # Remove entries with NA steps
  group_by(interval) %>% # Group into days
  summarise(meanSteps = mean(steps))  #Find mean steps each interval
```

Plot a time series plot of steps

```{r, echo=T, fig.height=6, fig.width=8}

ggplot() + geom_line(data = stepsPerInterval, aes(x = interval, y = meanSteps), binwidth = 1000 ) + theme(legend.position="none")

```

And find the maximum.

```{r, echo=T}
maxIntervalStart = stepsPerInterval$interval[which.max(stepsPerInterval$meanSteps)]
```

The interval with the most steps on average is from `r maxIntervalStart` to `r maxIntervalStart + 5` minutes.

## Imputing missing values

```{r, echo=T}
numberOfNas = sum(is.na(data$steps))
```
Of the `r nrow(data)` rows in the data, `r numberOfNas` are *NA* (`r round(100*numberOfNas / nrow(data), 2)`% of rows).

To replace these *NA*s, we will use the simple strategy of replacing them with the mean number of steps for that 5 minute interval, which we computed in the previous section. Then create a new data set replacing *NA*s with the mean for each interval.

```{r, echo=TRUE}
imputedData = data

stepsPerInterval = as.matrix(stepsPerInterval)
rownames(stepsPerInterval) = stepsPerInterval[,'interval']

imputedData$steps[is.na(imputedData$steps)] = stepsPerInterval[as.character(imputedData$interval[is.na(imputedData$steps)]),2]

```


## Are there differences in activity patterns between weekdays and weekends?

First we add the weekday/weekend factor and compute means by factor:
```{r, echo=TRUE}
imputedData$week = as.factor( 
  ifelse(
    weekdays(as.Date(imputedData$date)) %in% c("Saturday","Sunday"),
    "weekend", "weekday"
  ))

weekendMeans = imputedData %>% 
  filter(!is.na(steps)) %>% # Remove entries with NA steps
  group_by(week, interval) %>% # Group into weekend/day
  summarise(meanSteps = mean(steps))  #Find mean steps each interval
```

Then plot:
```{r, echo=TRUE}

ggplot() + geom_line(data = weekendMeans, aes(x = interval, y = meanSteps), binwidth = 1000 ) + facet_grid(week ~ .)

```
